---
title: "Plotting of heatmap of kinase activities with clustering rows and columns"
author: "Run Jin"
date: "11/15/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readr)
  library(pheatmap)
  library(readxl)
})
```

#### Define Directories 
```{r directories}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "kinase_activity_corr")

input_dir <- file.path(analysis_dir, "input")

plots_dir <- file.path(analysis_dir, "plots")
if(!dir.exists(plots_dir)){
  dir.create(plots_dir, recursive=TRUE)
}
```

#### Read in files necessary for analyses
```{r read files, echo=FALSE}
# histology file 
histology_df <- readr::read_tsv(file.path(data_dir, "pbta-histologies.tsv"), guess_max=10000)

# kinase activity score
kinase_activity <- readxl::read_excel(file.path(input_dir, "1-s2.0-S0092867420314513-mmc4.xlsx"), sheet =3, skip=2)
```

## Analyze HGAT Midline specifici PSI

```{r}
# PSI group 
psi_group <- readr::read_tsv(file.path(input_dir, "CLK1_PSI_grouping.txt"))

# match BS_ID to sample ID
histology_matched <- histology_df %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% psi_group$BS_id) %>%
  dplyr::select(Kids_First_Biospecimen_ID, sample_id, harmonized_diagnosis)
```

### what samples in PSI group-score file that are not in histologies file
```{r}
no_hist <- psi_group$BS_id[!psi_group$BS_id %in% histology_df$Kids_First_Biospecimen_ID]
no_hist
```

### what samples have both PSI score and kinase actvitiy scores
```{r}
# checkout common samples
common_samples <- histology_matched$sample_id[histology_matched$sample_id %in% colnames(kinase_activity)] %>%
  unique()
common_samples

```

### Prepare for heatmap plotting
```{r}
# generate an annotation file for the heatmap  
histology_matched_filtered <- histology_matched %>% 
  dplyr::filter(sample_id %in% common_samples) %>% 
  left_join( psi_group, by=c("Kids_First_Biospecimen_ID"= "BS_id")) %>%
  dplyr::distinct(sample_id, .keep_all = TRUE) %>%
  dplyr::select(-"Kids_First_Biospecimen_ID") %>% 
  tibble::column_to_rownames("sample_id") %>%
  arrange(Group, harmonized_diagnosis)

# filter kinase activity
kinase_activity_filtered <- kinase_activity %>% 
  tibble::column_to_rownames("proID") 

# clear the NA in the numbers
fix_NA_function <- function(df){
  df_fixed <- apply(df, 2, function(x) as.numeric(gsub("NA", "", x)))
  rownames(df_fixed) <- rownames(df)
  return(df_fixed)
}

kinase_activity_fixed <-fix_NA_function(kinase_activity_filtered) %>% 
  as.data.frame() %>% 
  # order the columns for plotting
  dplyr::select(rownames(histology_matched_filtered)) 

```

### save heatmap
```{r}
pheatmap::pheatmap(as.matrix(kinase_activity_fixed),
                   annotation_col = histology_matched_filtered,
                   cluster_rows=FALSE,
                   cluster_cols=FALSE,
                   width = 10,
                   height = 8,
                   filename = file.path(plots_dir, "kinase_group_psi_hist.pdf"))
```

### generate heatmap
```{r}
pheatmap::pheatmap(as.matrix(kinase_activity_fixed),
                   annotation_col = histology_matched_filtered,
                   cluster_rows=FALSE, 
                   cluster_cols=FALSE
                   )
```


## Analyze all Samples with PSI scores
```{r}
# PSI group 
psi_group <- readr::read_tsv(file.path(input_dir, "CLK1_PSI_grouping.txt"))

# match BS_ID to sample ID
histology_matched <- histology_df %>% 
  dplyr::filter(Kids_First_Biospecimen_ID %in% psi_group$BS_id) %>%
  dplyr::select(Kids_First_Biospecimen_ID, sample_id, harmonized_diagnosis)
```

### what samples in PSI group-score file that are not in histologies file
```{r}
no_hist <- psi_group$BS_id[!psi_group$BS_id %in% histology_df$Kids_First_Biospecimen_ID]
no_hist
```

### what samples have both PSI score and kinase actvitiy scores
```{r}
# checkout common samples
common_samples <- histology_matched$sample_id[histology_matched$sample_id %in% colnames(kinase_activity)] %>%
  unique()
common_samples

```

### Prepare for heatmap plotting
```{r}
# generate an annotation file for the heatmap  
histology_matched_filtered <- histology_matched %>% 
  dplyr::filter(sample_id %in% common_samples) %>% 
  left_join( psi_group, by=c("Kids_First_Biospecimen_ID"= "BS_id")) %>%
  dplyr::distinct(sample_id, .keep_all = TRUE) %>%
  dplyr::select(-"Kids_First_Biospecimen_ID") %>% 
  tibble::column_to_rownames("sample_id") %>%
  arrange(Group, harmonized_diagnosis)

# filter kinase activity
kinase_activity_filtered <- kinase_activity %>% 
  tibble::column_to_rownames("proID") 

# clear the NA in the numbers
fix_NA_function <- function(df){
  df_fixed <- apply(df, 2, function(x) as.numeric(gsub("NA", "", x)))
  rownames(df_fixed) <- rownames(df)
  return(df_fixed)
}

kinase_activity_fixed <-fix_NA_function(kinase_activity_filtered) %>% 
  as.data.frame() %>% 
  # order the columns for plotting
  dplyr::select(rownames(histology_matched_filtered)) 

```

### save heatmap
```{r}
pheatmap::pheatmap(as.matrix(kinase_activity_fixed),
                   annotation_col = histology_matched_filtered,
                   cluster_rows=FALSE,
                   cluster_cols=FALSE,
                   width = 10,
                   height = 8,
                   filename = file.path(plots_dir, "kinase_group_psi_hist.pdf"))
```

### generate heatmap
```{r}
pheatmap::pheatmap(as.matrix(kinase_activity_fixed),
                   annotation_col = histology_matched_filtered,
                   cluster_rows=FALSE, 
                   cluster_cols=FALSE
                   )
```
