---
title: "Run survival analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

**Purpose:** 

Runs survival analysis models in histology groups with splicing burden index (SBI) as a predictor

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Load packages, set directory paths and call setup script

```{r}
library(tidyverse)
library(survival)
library(ggpubr)
library(ggplot2)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
setwd(root_dir)

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
input_dir <- file.path(analysis_dir, "results")
results_dir <- file.path(analysis_dir, "results")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

source(file.path(analysis_dir, "util", "survival_models.R"))
```

Set metadata path

```{r set paths}
metadata_file <- file.path(input_dir, "splicing_indices_with_survival.tsv")
```

Read in metadata and consolidate HGG, LGG, and GNG subtypes into broader groups

```{r}
metadata <- read_tsv(metadata_file)

metadata <- metadata %>%
#  filter(!grepl("To be classified", molecular_subtype)) %>%
  mutate(mol_sub_group = case_when(
    Histology == "HGG" & grepl("wildtype", molecular_subtype) ~ "HGG, H3 WT",
    Histology == "HGG" & grepl("K28", molecular_subtype) ~ "DMG, H3 K28",
    grepl("G35", molecular_subtype) ~ "DHG, H3 G35",
    grepl("IHG", molecular_subtype) ~ "IHG",
    Histology == "HGG" & grepl("IDH", molecular_subtype) ~ "HGG, IDH",
    Histology == "LGG" & grepl("SEGA", molecular_subtype) ~ "SEGA",
    Histology == "LGG" & grepl("-BRAF", molecular_subtype) ~ "LGG, BRAF fusion",
    Histology == "LGG" & grepl("wildtype", molecular_subtype) ~ "LGG, WT",
    Histology == "LGG" & grepl("To be classified", molecular_subtype) ~ "LGG, To be classified",
    Histology == "LGG" ~ "LGG, Other alteration",
    Histology == "GNG" & grepl("wildtype", molecular_subtype) ~ "GNG/GNT, WT",
    Histology == "GNG" & grepl("-BRAF", molecular_subtype) ~ "GNG/GNT, BRAF fusion",
    Histology == "GNG" & !grepl("To be classified", molecular_subtype) ~ "GNG/GNT, Other alteration",
    !is.na(molecular_subtype) ~ molecular_subtype
  )) %>%
  dplyr::mutate(age_at_diagnosis_years = age_at_diagnosis_days/365.25)
```

Define histology groups to run survival analyses on 

```{r}
groups <- c("ATRT", "CPG", "EPN", "GNG", 
            "HGG", "LGG", "MB")

# create names vector for output files
plot_names <- c("atrt", "cpg", "epn", 
                "gng", "hgg", "lgg", "mb")
names(plot_names) <- groups
```

Loop through histology groups and perform the following: 

- Define samples with high and low SI burden (upper and lower quartiles, respectively)
- Run Kaplan-Meier survival analysis and save output for plotting
- Run cox proportional hazard models, including SI burden as covariate along with subtype (when applicable) and extent of tumor resection (LGG and GNG only). Save output for plotting

```{r}
for (group in groups){
  
  # define histology-specific results directory
  
  results_dir <- file.path(analysis_dir, "results", plot_names[group])
    
  if (!dir.exists(results_dir)) {
      dir.create(results_dir)
      
    }
  
  # Define output files for KM and coxph models
  km_os_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{plot_names[group]}_OS_SIburden.RDS"))
  
  coxph_os_model_file <- ifelse(group %in% c("GNG", "LGG"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_resection_SIburden.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_SIburden.RDS"))
  )
  
  km_efs_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{plot_names[group]}_EFS_SIburden.RDS"))
  
  coxph_efs_model_file <- ifelse(group %in% c("GNG", "LGG"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_resection_SIburden.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_SIburden.RDS"))
  )
  
  SI_burden_output_file <- file.path(results_dir,
                                     glue::glue("{plot_names[group]}_with_SI_group.tsv"))
  
  # Subset histology
  group_hist <- metadata %>%
    filter(grepl(group, Histology))
  
  # consolidate HGG subtypes into broader groups
  if (group == "HGG"){
    group_hist <- group_hist %>%
     dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("HGG, H3 WT",
                                                     "DHG, H3 G35",
                                                     "DMG, H3 K28",
                                                     "HGG, IDH",
                                                     "IHG"))) %>%
      arrange(mol_sub_group) %>%
      dplyr::mutate(tp53_status = case_when(
        grepl("TP53", molecular_subtype) ~ "TP53 altered",
        TRUE ~ "TP53 WT"
      )) %>%
      dplyr::mutate(tp53_status = fct_relevel(tp53_status,
                                              "TP53 WT", "TP53 altered"))
  }
    
  if (group %in% c("GNG", "LGG")){
    group_hist <- group_hist %>%
      dplyr::filter(!is.na(extent_of_tumor_resection) & !extent_of_tumor_resection %in% c("Not Applicable", "Not Reported", "Unavailable"))
  }

  #Define high and low SI burden samples using quartiles
  group_hist <- group_hist %>%
    dplyr::filter(!grepl("To be classified", molecular_subtype)) %>%
    dplyr::mutate(SI_group = case_when(
    SI > summary(SI)["3rd Qu."] ~ "High SI",
    SI < summary(SI)["1st Qu."] ~ "Low SI",
    TRUE ~ NA_character_
  )) %>%
    dplyr::mutate(SI_group = factor(SI_group,
                                    c("Low SI", "High SI")))
  
  # Generate kaplan meier survival models for OS and PFS, and save outputs
  kap_os <- survival_analysis(
  metadata  = group_hist,
  ind_var = "SI_group",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "OS_days",
  status_col = "OS_status"
  )

  readr::write_rds(kap_os, km_os_model_file)
  
  kap_efs <- survival_analysis(
  metadata  = group_hist,
  ind_var = "SI_group",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "EFS_days",
  status_col = "EFS_status"
  )

  readr::write_rds(kap_efs, km_efs_model_file)

  # Run cox proprtional hazards models for OS, including appropriate covariates per histology
  add_model_os <- suppressWarnings(fit_save_model(group_hist,
                                  terms = ifelse(
                                      group %in% c("GNG", "LGG"), "mol_sub_group+extent_of_tumor_resection+age_at_diagnosis_years+SI_group",
                                      ifelse(
                                        group == "HGG",
                                        "mol_sub_group+tp53_status+age_at_diagnosis_years+SI_group",
                                        ifelse(
                                          group %in% c("ATRT", "EPN", "MB"), 
                                          "mol_sub_group+age_at_diagnosis_years+SI_group",
                                          "age_at_diagnosis_years+SI_group"
                                  ))),
             coxph_os_model_file,
             "multivariate",
             years_col = "OS_years",
             status_col = "OS_status"
            ))
  
  # Run cox proprtional hazards models for PFS, including appropriate covariates per histology
  add_model_efs <- suppressWarnings(fit_save_model(group_hist,
                                  terms = ifelse(
                                      group %in% c("GNG", "LGG"), "mol_sub_group+extent_of_tumor_resection+age_at_diagnosis_years+SI_group",
                                      ifelse(
                                        group == "HGG",
                                        "mol_sub_group+tp53_status+age_at_diagnosis_years+SI_group",
                                        ifelse(
                                          group %in% c("ATRT", "EPN", "MB"), 
                                          "mol_sub_group+age_at_diagnosis_years+SI_group",
                                          "age_at_diagnosis_years+SI_group"
                                  ))),
             coxph_efs_model_file,
             "multivariate",
             years_col = "EFS_years",
             status_col = "EFS_status"
            ))
  
  group_hist %>%
    filter(!is.na(SI_group)) %>%
    write_tsv(SI_burden_output_file)
  
  
}

```

Run same survival analyses on molecular subtypes with sufficient sample size

```{r}
# define subtype df with histology, subtype, and file names
subtype_df <- data.frame(hist = c("CPG", "EPN",
                                  rep("HGG", 2), "GNG",
                                  rep("LGG", 3), rep("MB", 3)),
                         names = c("cpg", "epn",
                                  rep("hgg", 2), "gng",
                                  rep("lgg", 3), rep("mb", 3)),
                         subtype = c("ADAM", "PF A", "K28", "WT",
                                     "Other", "BRAF", "Other", "WT", 
                                     "Group3", "Group4", "SHH"))

# Loop through subtypes and run survival analyses
for (i in 1:nrow(subtype_df)) {
  
  # define histology-specific results dir
  results_dir <- file.path(analysis_dir, "results", subtype_df$names[i])
    
  if (!dir.exists(results_dir)) {
      dir.create(results_dir)
      
    }
  
  # define output files
  
  km_os_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{subtype_df$names[i]}_{subtype_df$subtype[i]}_OS_SIburden.RDS"))
  
  coxph_os_model_file <- ifelse(subtype_df$hist[i] %in% c("LGG", "GNG"),
                             file.path(results_dir,
                                      glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_OS_additive_terms_subtype_resection_SIburden.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_OS_additive_terms_subtype_SIburden.RDS"))
  )
  
  km_efs_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{subtype_df$names[i]}_{subtype_df$subtype[i]}_EFS_SIburden.RDS"))
  
  coxph_efs_model_file <- ifelse(subtype_df$hist[i] %in% c("LGG", "GNG"),
                             file.path(results_dir,
                                       glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_EFS_additive_terms_subtype_resection_SIburden.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype[i]}_EFS_additive_terms_subtype_SIburden.RDS"))
  )
  
  SI_burden_output_file <- file.path(results_dir,
                                     glue::glue("{subtype_df$names[i]}_{subtype_df$subtype[i]}_with_SI_group.tsv"))
  
    #Subset metadata for histology subtype
    subtype_hist <- metadata %>%
    dplyr::filter(Histology == subtype_df$hist[i] & grepl(subtype_df$subtype[i], mol_sub_group)) %>%
      mutate(SI_group = case_when(
      SI > summary(SI)["3rd Qu."] ~ "High SI",
      SI < summary(SI)["1st Qu."] ~ "Low SI",
      TRUE ~ NA_character_
    ))%>%
    dplyr::mutate(SI_group = factor(SI_group,
                                    c("Low SI", "High SI")))
    
    if (subtype_df$hist[i] == "HGG"){
      subtype_hist <- subtype_hist %>%
        dplyr::mutate(tp53_status = case_when(
          grepl("TP53", molecular_subtype) ~ "TP53 altered",
          TRUE ~ "TP53 WT"
        )) %>%
        dplyr::mutate(tp53_status = fct_relevel(tp53_status,
                                                "TP53 WT", "TP53 altered"))
    }
    
    if (subtype_df$hist[i] %in% c("GNG", "LGG")){
      subtype_hist <- subtype_hist %>%
        dplyr::filter(!is.na(extent_of_tumor_resection) & !extent_of_tumor_resection %in% c("Not Applicable", "Not Reported", "Unavailable"))
    }
    
    #Run kaplan meier survival models and save to output
      kap_os <- survival_analysis(
                  metadata  = subtype_hist,
                  ind_var = "SI_group",
                  test = "kap.meier",
                  metadata_sample_col = "Kids_First_Biospecimen_ID",
                  days_col = "OS_days",
                  status_col = "OS_status"
                  )

  readr::write_rds(kap_os, km_os_model_file)
  
  kap_efs <- survival_analysis(
                metadata  = subtype_hist,
                ind_var = "SI_group",
                test = "kap.meier",
                metadata_sample_col = "Kids_First_Biospecimen_ID",
                days_col = "EFS_days",
                status_col = "EFS_status"
                )
  
    readr::write_rds(kap_efs, km_efs_model_file)
    
    # Run coxph models and save to output
    add_model_os <- suppressWarnings(fit_save_model(subtype_hist,
                                  terms = ifelse(subtype_df$hist[i] %in% c("GNG", "LGG"),
                                                 "extent_of_tumor_resection+age_at_diagnosis_years+SI_group",
                                                 ifelse(subtype_df$hist[i] == "HGG",
                                                        "tp53_status+age_at_diagnosis_years+SI_group",
                                                        "age_at_diagnosis_years+SI_group")),
                                   coxph_os_model_file,
                                   "multivariate",
                                   years_col = "OS_years",
                                   status_col = "OS_status"))
    
    add_model_efs <- suppressWarnings(fit_save_model(subtype_hist,
                                  terms = ifelse(subtype_df$hist[i] %in% c("GNG", "LGG"),
                                                 "extent_of_tumor_resection+age_at_diagnosis_years+SI_group",
                                                 ifelse(subtype_df$hist[i] == "HGG",
                                                        "tp53_status+age_at_diagnosis_years+SI_group",
                                                        "age_at_diagnosis_years+SI_group")),
                                     coxph_efs_model_file,
                                     "multivariate",
                                     years_col = "EFS_years",
                                     status_col = "EFS_status"
                                    ))
  subtype_hist %>%
    filter(!is.na(SI_group)) %>%
    write_tsv(SI_burden_output_file)
  
    
}

```