---
title: "Assess survival in patients with HGG by CLK1 status"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett, Jo Lynne Rokita
date: 2023
params:
  plot_ci: TRUE
---

**Purpose:** 

- Generate survival models with CLK1 status (high vs. low) as covariate
- Plot survival by CLK1 status

## Setup

#### Packages and functions

load libraries and set directories

```{r Set up library}
library(survival)
library(ggpubr)
library(tidyverse)
library('data.table')
library(patchwork)

# Set directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Load survival functions
source(file.path(analysis_dir, "util", "survival_models.R"))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

Set file paths

```{r}
histologies_file <- file.path(results_dir, "splicing_indices_with_survival.tsv")

diff_psi_file <- file.path(data_dir, "splice-events-rmats.tsv.gz")
```

Load histologies and subset for HGG and DMG

```{r}
hgg_hist <- read_tsv(histologies_file) %>%
  dplyr::filter(Histology %in% c("DIPG or DMG", "Other high-grade glioma")) %>%
  dplyr::select(Kids_First_Biospecimen_ID, mol_sub_group,
                OS_days, OS_years, OS_status,
                EFS_days, EFS_years, EFS_status)
```

Load rMATs file and subset for CLK1 exon 4 PSI values

```{r}
splice_CLK1_df <-  fread(diff_psi_file) %>%
   dplyr::filter(geneSymbol=="CLK1",
                 exonStart_0base=="200860124", exonEnd=="200860215") %>%
  dplyr::rename('Kids_First_Biospecimen_ID'=sample_id,
                PSI = IncLevel1) %>%
  dplyr::select(Kids_First_Biospecimen_ID, PSI)

# join PSI value to hgg_hist
hgg_hist <- hgg_hist %>%
  left_join(splice_CLK1_df)
```

Calculate CLK1 quantiles and define samples as Low, Medium, High

```{r}
quantiles_clk1 <- quantile(hgg_hist$PSI, probs=c(.25, .75), na.rm = TRUE)
lower_sbi <- quantiles_clk1[1]
upper_sbi <- quantiles_clk1[2]

hgg_hist <- hgg_hist %>%
  mutate(clk1_status = case_when(PSI > upper_sbi ~ "CLK1 exon 4 High",
                                 PSI < lower_sbi ~ "CLK1 exon 4 Low",
                                 TRUE ~ "Middle"))
```

Create survival df with only high & low CLK1 status, and relevel mol subgroups

```{r}
survival_df <- hgg_hist %>%
  dplyr::filter(clk1_status %in% c("CLK1 exon 4 High", "CLK1 exon 4 Low")) %>%
  dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                            c("HGG, H3 wildtype", "HGG, H3 wildtype, TP53",
                                              "DHG, H3 G35", "DHG, H3 G35, TP53",
                                              "DMG, H3 K28", "DMG, H3 K28, TP53",
                                              "HGG, IDH, TP53", "HGG, PXA",
                                              "HGG, PXA, TP53", "IHG, NTRK-altered",
                                              "IHG, ROS1-altered")))
```

Generate KM models wiht `clk1_status` as variable

```{r}
kap_os <- survival_analysis(
  metadata  = survival_df,
  ind_var = "clk1_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "OS_days",
  status_col = "OS_status"
)

readr::write_rds(kap_os, file.path(results_dir,
                                   "logrank_HGG_OS_CLK1_status.RDS"))

kap_efs <- survival_analysis(
  metadata  = survival_df,
  ind_var = "clk1_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "EFS_days",
  status_col = "EFS_status"
)

readr::write_rds(kap_efs, file.path(results_dir,
                                    "logrank_HGG_EFS_CLK1_status.RDS"))
```

Create KM plot

```{r}
km_plot <- plotKM(model = list(kap_os, kap_efs),
                  variable = "clk1_status",
                  combined = T, 
                  title = "")
              
ggsave(file.path(plot_dir, "km_HGG_DMG_OS_EFS_CLK1_status.pdf"), km_plot,
       width = 10, height = 6, units = "in",
       device = "pdf")
```

### Generate cox proportional hazards models 

Interaction model OS:

```{r}
int_os_surv_result <- file.path(results_dir, 
                     "cox_HGG_OS_interaction_terms_subtype_clk1_status.RDS")

int_model_os <- fit_save_model(survival_df,
                               terms = "mol_sub_group*clk1_status",
                               int_os_surv_result,
                               "multivariate",
                               years_col = "OS_years",
                               status_col = "OS_status"
                              )

int_os_forest_plot <- plotForest(readRDS(int_os_surv_result))

ggsave(file.path(plot_dir, "forest_HGG_DMG_OS_int_subtype_clk1_status.pdf"), 
       int_os_forest_plot, width = 12, height = 3)
```

OS additive model: 

```{r}
add_os_surv_result <- file.path(results_dir, 
                     "cox_HGG_OS_additive_terms_subtype_clk1_status.RDS")

add_model_os <- fit_save_model(survival_df,
                               terms = "mol_sub_group+clk1_status",
                               add_os_surv_result,
                               "multivariate",
                               years_col = "OS_years",
                               status_col = "OS_status"
                              )

add_os_forest_plot <- plotForest(readRDS(add_os_surv_result))

ggsave(file.path(plot_dir, "forest_HGG_DMG_OS_add_subtype_clk1_status.pdf"), 
       add_os_forest_plot, width = 7.5, height = 3)
```

EFS interaction model:

```{r}
int_efs_surv_result <- file.path(results_dir, 
                     "cox_HGG_EFS_interaction_terms_subtype_clk1_status.RDS")

int_model_efs <- fit_save_model(survival_df,
                               terms = "mol_sub_group*clk1_status",
                               int_efs_surv_result,
                               "multivariate",
                               years_col = "EFS_years",
                               status_col = "EFS_status"
                              )

int_efs_forest_plot <- plotForest(readRDS(int_efs_surv_result))

ggsave(file.path(plot_dir, "forest_HGG_DMG_EFS_int_subtype_clk1_status.pdf"), 
       int_efs_forest_plot, width = 12, height = 4)
```

EFS additive model: 

```{r}
add_efs_surv_result <- file.path(results_dir, 
                     "cox_HGG_EFS_additive_terms_subtype_clk1_status.RDS")

add_model_efs <- fit_save_model(survival_df,
                               terms = "mol_sub_group+clk1_status",
                               add_efs_surv_result,
                               "multivariate",
                               years_col = "EFS_years",
                               status_col = "EFS_status"
                              )

add_efs_forest_plot <- plotForest(readRDS(add_efs_surv_result))

ggsave(file.path(plot_dir, "forest_HGG_EFS_add_subtype_clk1_status.pdf"), 
       add_efs_forest_plot, width = 7.5, height = 3)
```

There appears to be a subtype x clk1_status interaction effect, such that survival trends may differ in DMG vs. other HGG. 

Create models and plots for DMG separately:

```{r}
kap_os_dmg <- survival_analysis(
  metadata  = survival_df[grepl("DMG", survival_df$mol_sub_group),],
  ind_var = "clk1_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "OS_days",
  status_col = "OS_status"
)

readr::write_rds(kap_os_dmg, file.path(results_dir,
                                       "logrank_DMG_OS_CLK1_status.RDS"))

kap_efs_dmg <- survival_analysis(
  metadata  = survival_df[grepl("DMG", survival_df$mol_sub_group),],
  ind_var = "clk1_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "EFS_days",
  status_col = "EFS_status"
)

readr::write_rds(kap_efs_dmg, file.path(results_dir,
                                        "logrank_DMG_EFS_CLK1_status.RDS"))
```

Plot KM

```{r}
km_plot <- plotKM(model = list(kap_os_dmg, kap_efs_dmg),
                  variable = "clk1_status",
                  combined = T, 
                  title = "")
              
ggsave(file.path(plot_dir, "km_DMG_OS_EFS_CLK1_status.pdf"), km_plot,
       width = 10, height = 6, units = "in",
       device = "pdf")
```

Generate and plot DMG survival models:

```{r}
survival_dmg_df <- survival_df %>%
  dplyr::filter(grepl("DMG", mol_sub_group)) %>%
  dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                            c("DMG, H3 K28",
                                              "DMG, H3 K28, TP53")))

add_dmg_os_surv_result <- file.path(results_dir, 
                     "cox_DMG_OS_additive_terms_subtype_clk1_status.RDS")

add_model_os_dmg <- fit_save_model(survival_dmg_df,
                               terms = "mol_sub_group+clk1_status",
                               add_dmg_os_surv_result,
                               "multivariate",
                               years_col = "OS_years",
                               status_col = "OS_status"
                              )

add_os_dmg_forest_plot <- plotForest(readRDS(add_dmg_os_surv_result))

ggsave(file.path(plot_dir, "forest_DMG_OS_add_subtype_clk1_status.pdf"), 
       add_os_dmg_forest_plot, width = 7.5, height = 3)
```

```{r}
add_dmg_efs_surv_result <- file.path(results_dir, 
                     "cox_DMG_EFS_additive_terms_subtype_clk1_status.RDS")

add_model_efs_dmg <- fit_save_model(survival_dmg_df,
                               terms = "mol_sub_group+clk1_status",
                               add_dmg_efs_surv_result,
                               "multivariate",
                               years_col = "EFS_years",
                               status_col = "EFS_status"
                              )

add_efs_dmg_forest_plot <- plotForest(readRDS(add_dmg_efs_surv_result))

ggsave(file.path(plot_dir, "forest_DMG_EFS_add_subtype_clk1_status.pdf"), 
       add_efs_dmg_forest_plot, width = 7.5, height = 3)
```

Generate and plot models for HGG separately: 

```{r}
kap_os_hgg <- survival_analysis(
  metadata  = survival_df[!grepl("DMG", survival_df$mol_sub_group),],
  ind_var = "clk1_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "OS_days",
  status_col = "OS_status"
)

readr::write_rds(kap_os_hgg, file.path(results_dir,
                                       "logrank_HGG_OS_CLK1_status.RDS"))

kap_efs_hgg <- survival_analysis(
  metadata  = survival_df[!grepl("DMG", survival_df$mol_sub_group),],
  ind_var = "clk1_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID",
  days_col = "EFS_days",
  status_col = "EFS_status"
)

readr::write_rds(kap_efs_hgg, file.path(results_dir,
                                        "logrank_HGG_EFS_CLK1_status.RDS"))
```

Plot KM 

```{r}
km_plot <- plotKM(model = list(kap_os_hgg, kap_efs_hgg),
                  variable = "clk1_status",
                  combined = T, 
                  title = "")
              
ggsave(file.path(plot_dir, "km_HGG_OS_EFS_CLK1_status.pdf"), km_plot,
       width = 10, height = 6, units = "in",
       device = "pdf")
```

Generate and plot coxph models in HGG:  

```{r}
add_hgg_os_surv_result <- file.path(results_dir, 
                     "cox_HGG_OS_additive_terms_subtype_clk1_status.RDS")

add_model_os_hgg <- fit_save_model(survival_df[!grepl("DMG", survival_df$mol_sub_group),],
                               terms = "mol_sub_group+clk1_status",
                               add_hgg_os_surv_result,
                               "multivariate",
                               years_col = "OS_years",
                               status_col = "OS_status"
                              )

add_os_hgg_forest_plot <- plotForest(readRDS(add_hgg_os_surv_result))

ggsave(file.path(plot_dir, "forest_HGG_OS_add_subtype_clk1_status.pdf"), 
       add_os_hgg_forest_plot, width = 7.5, height = 3)
```

```{r}
add_hgg_efs_surv_result <- file.path(results_dir, 
                     "cox_HGG_EFS_additive_terms_subtype_clk1_status.RDS")

add_model_efs_hgg <- fit_save_model(survival_df[!grepl("DMG", survival_df$mol_sub_group),],
                               terms = "mol_sub_group+clk1_status",
                               add_hgg_efs_surv_result,
                               "multivariate",
                               years_col = "EFS_years",
                               status_col = "EFS_status"
                              )

add_efs_hgg_forest_plot <- plotForest(readRDS(add_hgg_efs_surv_result))

ggsave(file.path(plot_dir, "forest_HGG_EFS_add_subtype_clk1_status.pdf"), 
       add_efs_hgg_forest_plot, width = 7.5, height = 3)
```

Print session info

```{r}
sessionInfo()
```
