---
title: "Run NF1-215 PSI survival analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett, Jo Lynne Rokita
date: 2023, 2024
params:
  plot_ci: TRUE
---

**Purpose:** 

Runs survival analysis models in histology groups with NF1-215 PSI as a predictor

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Load packages, set directory paths and call setup script

```{r}
library(tidyverse)
library(survival)
library(ggpubr)
library(ggplot2)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
input_dir <- file.path(analysis_dir, "results")
results_dir <- file.path(analysis_dir, "results")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

source(file.path(analysis_dir, "util", "survival_models.R"))
```

Set metadata path

```{r set paths}
metadata_file <- file.path(input_dir, "splicing_indices_with_survival.tsv")

nf1_psi_file <- file.path(root_dir, "analyses",
                          "CLK1-splicing_correlations", "results",
                          "clk1-nf1-psi-exp-df.rds")
```

Wrangle data
Append `NF1-215-PSI` to `metadata`

```{r}
metadata <- read_tsv(metadata_file)

nf1_psi <- readRDS(nf1_psi_file) %>%
  rownames_to_column("Kids_First_Biospecimen_ID")

metadata <- metadata %>%
  left_join(nf1_psi %>% dplyr::select(Kids_First_Biospecimen_ID,
                                      `NF1-215_PSI`))
  

```

Define histology groups to run survival analyses on 

```{r}

subtype_df <- read_tsv(file.path(analysis_dir, "results", "subtypes-for-survival.tsv"))

groups <- unique(subtype_df$hist)

# folder names
folder <- unique(subtype_df$foldername)
names(folder) <- groups
  
# create names vector for output files
plot_names <- unique(subtype_df$names)
names(plot_names) <- groups

```

Loop through histology groups and perform the following: 

- Define samples with high and low NF1-215 PSI (upper and lower quartiles, respectively)
- Run Kaplan-Meier survival analysis and save output for plotting
- Run cox proportional hazard models, including SI burden as covariate along with subtype (when applicable) and extent of tumor resection (LGG and GNG only). Save output for plotting

```{r}
for (group in groups){
  
  results_dir <- file.path(analysis_dir, "results", folder[[group]])
  
  if (!dir.exists(results_dir)) {
    dir.create(results_dir)
      
    }
  
  # Define output files for KM and coxph models
  km_os_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{plot_names[group]}_OS_nf1_psi_group.RDS"))
  
  coxph_os_model_file <- ifelse(group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_resection_nf1_psi_group.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_nf1_psi_group.RDS"))
  )
  
  km_efs_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{plot_names[group]}_EFS_nf1_psi_group.RDS"))
  
  coxph_efs_model_file <- ifelse(group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_resection_nf1_psi_group.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_nf1_psi_group.RDS"))
  )
  
  nf1_psi_output_file <- file.path(results_dir,
                                     glue::glue("{plot_names[group]}_with_nf1_215_psi.tsv"))
  
  # Subset histology
  group_hist <- metadata %>%
    filter(grepl(group, Histology))
  
  if (group == "DIPG or DMG"){
    group_hist <- group_hist %>%
     dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("HGG, H3 wildtype", "HGG, H3 wildtype, TP53",
                                                                "DMG, H3 K28", "DMG, H3 K28, TP53"
                                                                )))
  }  
  
  # consolidate HGG subtypes into broader groups
  if (group == "Other high-grade glioma"){
    group_hist <- group_hist %>%
     dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("HGG, H3 wildtype", "HGG, H3 wildtype, TP53",
                                                                "DHG, H3 G35", "DHG, H3 G35, TP53",
                                                                "DMG, H3 K28", "DMG, H3 K28, TP53",
                                                                "HGG, IDH, TP53", "HGG, PXA", "HGG, PXA, TP53", 
                                                                "IHG, ALK-altered", "IHG, NTRK-altered",
                                                                "IHG, NTRK-altered, TP53", "IHG, ROS1-altered"
                                                                )))
  }
  
  
  if (group == "Low-grade glioma"){
    group_hist <- group_hist %>%
     dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("Wildtype", "BRAF V600E", "BRAF fusion",
                                                                "Other alteration", "SEGA"
                                                                )))
  }
  
  if (group == "Ependymoma"){
    group_hist <- group_hist %>%
     dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("EPN, PF A", "EPN, PF B",
                                                                "EPN, ST ZFTA", "EPN, ST YAP1",
                                                                "EPN, MPE", "EPN, SP", "EPN, SP-MYCN"
                                                                ))) 
  }
  
    if (group == "Mixed neuronal-glial tumor"){
    group_hist <- group_hist %>%
     dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("Wildtype", "BRAF V600E", "BRAF fusion",
                                                                "Other alteration"
                                                                )))
  }
  
  #Define high and low NF1-215 PSI samples using quartiles
  
  if (group %in% c("Low-grade glioma", "Mixed neuronal-glial tumor")){
    
    group_hist <- group_hist %>%
      filter(!is.na(mol_sub_group) & !is.na(OS_days) & !is.na(EFS_days) & !is.na(extent_of_tumor_resection) & !extent_of_tumor_resection %in% c("Unavailable", "Not Applicable")) %>%
      dplyr::mutate(NF1_psi_group = case_when(
      `NF1-215_PSI` > summary(`NF1-215_PSI`)["3rd Qu."] ~ "High NF1-215 PSI",
      `NF1-215_PSI` < summary(`NF1-215_PSI`)["1st Qu."] ~ "Low NF1-215 PSI",
      TRUE ~ NA_character_
    )) %>%
      dplyr::mutate(NF1_psi_group = factor(NF1_psi_group,
                                      c("Low NF1-215 PSI", "High NF1-215 PSI")))
    
  } else {
    group_hist <- group_hist %>%
      filter(!is.na(mol_sub_group) & !is.na(OS_days) & !is.na(EFS_days)) %>%
      dplyr::mutate(NF1_psi_group = case_when(
      `NF1-215_PSI` > summary(`NF1-215_PSI`)["3rd Qu."] ~ "High NF1-215 PSI",
      `NF1-215_PSI` < summary(`NF1-215_PSI`)["1st Qu."] ~ "Low NF1-215 PSI",
      TRUE ~ NA_character_
    )) %>%
      dplyr::mutate(NF1_psi_group = factor(NF1_psi_group,
                                      c("Low NF1-215 PSI", "High NF1-215 PSI")))
    
  }
  
  # Generate kaplan meier survival models for OS and EFS, and save outputs
  kap_os <- survival_analysis(
    metadata  = group_hist,
    ind_var = "NF1_psi_group",
    test = "kap.meier",
    metadata_sample_col = "Kids_First_Biospecimen_ID",
    days_col = "OS_days",
    status_col = "OS_status"
  )

  readr::write_rds(kap_os, km_os_model_file)
  
  kap_efs <- survival_analysis(
    metadata  = group_hist,
    ind_var = "NF1_psi_group",
    test = "kap.meier",
    metadata_sample_col = "Kids_First_Biospecimen_ID",
    days_col = "EFS_days",
    status_col = "EFS_status"
  )

  readr::write_rds(kap_efs, km_efs_model_file)

  # Run cox proprtional hazards models for OS, including appropriate covariates per histology
  add_model_os <- suppressWarnings(fit_save_model(group_hist,
                                  terms = ifelse(
                                      group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"), "mol_sub_group+extent_of_tumor_resection+NF1_psi_group",
                                      ifelse(group %in% c("Atypical Teratoid Rhabdoid Tumor", "Ependymoma", "DIPG or DMG", "Other high-grade glioma", "Medulloblastoma"), "mol_sub_group+NF1_psi_group", "NF1_psi_group")
                                      ),
             coxph_os_model_file,
             "multivariate",
             years_col = "OS_years",
             status_col = "OS_status"
            ))
  
  # Run cox proprtional hazards models for EFS, including appropriate covariates per histology
  add_model_efs <- suppressWarnings(fit_save_model(group_hist,
                                  terms = ifelse(
                                      group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"), "mol_sub_group+extent_of_tumor_resection+NF1_psi_group",
                                      ifelse(group %in% c("Atypical Teratoid Rhabdoid Tumor", "Ependymoma", "DIPG or DMG", "Other high-grade glioma", "Medulloblastoma"),
                                             "mol_sub_group+NF1_psi_group",
                                             "NF1_psi_group")
                                      ),
             coxph_efs_model_file,
             "multivariate",
             years_col = "EFS_years",
             status_col = "EFS_status"
            ))
  
  group_hist %>%
    filter(!is.na(NF1_psi_group)) %>%
    select(-OS_years, -EFS_years, 
           -Kids_First_Participant_ID) %>%
    write_tsv(nf1_psi_output_file)
  
}
```

Run same survival analyses on molecular subtypes with sufficient sample size

```{r}
# Loop through subtypes and run survival analyses
for (i in 1:nrow(subtype_df)) {
  
  results_dir <- file.path(analysis_dir, "results", subtype_df$foldername[i])
  
  # define output files
  km_os_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_OS_nf1_psi_group.RDS"))
  
  coxph_os_model_file <- ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"),
                             file.path(results_dir,
                                      glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_OS_additive_terms_subtype_resection_nf1_psi_group.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_OS_additive_terms_subtype_nf1_psi_group.RDS"))
  )
  
  km_efs_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_EFS_nf1_psi_group.RDS"))
  
  coxph_efs_model_file <- ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"),
                             file.path(results_dir,
                                       glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_EFS_additive_terms_subtype_resection_nf1_psi_group.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_EFS_additive_terms_subtype_nf1_psi_group.RDS"))
  )
  
  SI_burden_output_file <- file.path(results_dir,
                                     glue::glue("{subtype_df$names[i]}_{subtype_df$subtype_name[i]}_with_nf1_psi_group.tsv"))
  
    #Subset metadata for histology subtype
    subtype_hist <- metadata %>%
      filter(Histology == subtype_df$hist[i] & grepl(subtype_df$subtype[i], mol_sub_group))
    
    
    if(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma")){
      
      subtype_hist <- subtype_hist %>%
        dplyr::filter(!is.na(extent_of_tumor_resection) & !extent_of_tumor_resection %in% c("Not Applicable", "Unavailable")) %>%
        dplyr::mutate(NF1_psi_group = case_when(
        `NF1-215_PSI` > summary(`NF1-215_PSI`)["3rd Qu."] ~ "High NF1-215 PSI",
        `NF1-215_PSI` < summary(`NF1-215_PSI`)["1st Qu."] ~ "Low NF1-215 PSI",
        TRUE ~ NA_character_
      )) %>%
        dplyr::mutate(NF1_psi_group = factor(NF1_psi_group,
                                        c("Low NF1-215 PSI", "High NF1-215 PSI")))
      
    } else {
      
      subtype_hist <- subtype_hist %>%
      dplyr::mutate(NF1_psi_group = case_when(
        `NF1-215_PSI` > summary(`NF1-215_PSI`)["3rd Qu."] ~ "High NF1-215 PSI",
        `NF1-215_PSI` < summary(`NF1-215_PSI`)["1st Qu."] ~ "Low NF1-215 PSI",
        TRUE ~ NA_character_
      )) %>%
      dplyr::mutate(NF1_psi_group = factor(NF1_psi_group,
                                      c("Low NF1-215 PSI", "High NF1-215 PSI")))
      
    }
    
    #Run kaplan meier survival models and save to output
  kap_os <- survival_analysis(
              metadata  = subtype_hist,
              ind_var = "NF1_psi_group",
              test = "kap.meier",
              metadata_sample_col = "Kids_First_Biospecimen_ID",
              days_col = "OS_days",
              status_col = "OS_status"
              )

  readr::write_rds(kap_os, km_os_model_file)
  
  kap_efs <- survival_analysis(
                metadata  = subtype_hist,
                ind_var = "NF1_psi_group",
                test = "kap.meier",
                metadata_sample_col = "Kids_First_Biospecimen_ID",
                days_col = "EFS_days",
                status_col = "EFS_status"
                )
  
  readr::write_rds(kap_efs, km_efs_model_file)
    
    # Run coxph models and save to output
  add_model_os <- suppressWarnings(fit_save_model(subtype_hist,
                                terms = ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"),
                                               "extent_of_tumor_resection+NF1_psi_group",
                                               ifelse(subtype_df$hist[i] == "Other high-grade glioma", 
                                                      "mol_sub_group+NF1_psi_group",
                                                      "NF1_psi_group"
                                                      )),
                                 coxph_os_model_file,
                                 "multivariate",
                                 years_col = "OS_years",
                                 status_col = "OS_status")
                                )
  
  add_model_efs <- suppressWarnings(fit_save_model(subtype_hist,
                                 terms = ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma"),
                                               "extent_of_tumor_resection+NF1_psi_group",
                                               ifelse(subtype_df$hist[i] == "Other high-grade glioma", 
                                                      "mol_sub_group+NF1_psi_group",
                                                      "NF1_psi_group"
                                                      )),
                                   coxph_efs_model_file,
                                   "multivariate",
                                   years_col = "EFS_years",
                                   status_col = "EFS_status"
                                  ))
  subtype_hist %>%
    filter(!is.na(NF1_psi_group)) %>%
    select(-OS_years, -EFS_years,
           -Kids_First_Participant_ID) %>%
    write_tsv(SI_burden_output_file)
    
}
```