---
title: "CLK1 PSI-associated survival analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2024
params:
  plot_ci: TRUE
---

**Purpose:** 

- Generate survival models with CLK1 exon 4 PSI as covariate

## Setup

#### Packages and functions

load libraries and set directories

```{r Set up library}
library(survival)
library(ggpubr)
library(tidyverse)
library(data.table)
library(patchwork)

# Set directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Load survival functions
source(file.path(analysis_dir, "util", "survival_models.R"))

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

Set file paths

```{r}
histologies_file <- file.path(results_dir, "splicing_indices_with_survival.tsv")

diff_psi_file <- file.path(root_dir, "analyses",
                           "CLK1-splicing_correlations", "results",
                           "clk1-splice-events-rmats.tsv")
```

Load histologies

```{r}
hist <- read_tsv(histologies_file) %>%
  # there are several MB samples we have confirmed had partial resection; update these:
  dplyr::mutate(extent_of_tumor_resection = case_when(
    Kids_First_Participant_ID %in% c("PT_KNJDVV4K",
                                     "PT_M0AP3Z15",
                                     "PT_26V72ZXB") ~ "Partial resection",
    TRUE ~ extent_of_tumor_resection
    )) %>%
  dplyr::mutate(extent_of_tumor_resection = fct_relevel(extent_of_tumor_resection,
                                                      c("Gross/Near total resection",
                                                        "Partial resection",
                                                        "Biopsy only", "Not Reported",
                                                        "Unavailable")))
```

Load rMATs file and subset for CLK1 exon 4 PSI values

```{r}
splice_CLK1_df <-  fread(diff_psi_file) %>%
   dplyr::filter(geneSymbol=="CLK1",
                 exonStart_0base=="200860124", exonEnd=="200860215") %>%
  dplyr::rename('Kids_First_Biospecimen_ID'=sample_id,
                clk1_PSI = IncLevel1) %>%
  dplyr::select(Kids_First_Biospecimen_ID, clk1_PSI)

# join PSI value to hgg_hist
hist <- hist %>%
  left_join(splice_CLK1_df)
```

Define histology groups to run survival analyses on 

```{r}
subtype_df <- read_tsv(file.path(results_dir, "subtypes-for-survival.tsv"))

groups <- unique(subtype_df$hist)

# folder names
folder <- unique(subtype_df$foldername)
names(folder) <- groups
  
# create names vector for output files
plot_names <- unique(subtype_df$names)
names(plot_names) <- groups
```

Loop through histology groups and perform the following: 

- Filter hist by histology
- Run cox proportional hazard models, including CLK1 exon 4 PSI as covariate along with subtype (when applicable) and extent of tumor resection

```{r}
for (group in groups){
  
  results_dir <- file.path(analysis_dir, "results", folder[[group]])
  
  plot_dir <- file.path(analysis_dir, "plots", folder[[group]])

  coxph_os_model_file <- ifelse(group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                             "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                             "Medulloblastoma"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_resection_clk1_psi.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_clk1_psi.RDS"))
  )

  coxph_efs_model_file <- ifelse(group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                             "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                             "Medulloblastoma"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_resection_clk1_psi.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_clk1_psi.RDS"))
  )
  
    # Subset histology
  group_hist <- hist %>%
    filter(grepl(group, Histology))
  
  # Subset histology and transform SI 
  
    if (group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                     "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                     "Medulloblastoma")){
      
      group_hist <- group_hist %>%
        filter(!is.na(mol_sub_group) & !is.na(OS_days) & !is.na(EFS_days) & !is.na(extent_of_tumor_resection) & !extent_of_tumor_resection %in% c("Unavailable", "Not Reported")) %>%
        dplyr::mutate(clk1_PSI = clk1_PSI * 10)
      
    } else {
      
      group_hist <- group_hist %>%
        filter(!is.na(mol_sub_group) & !is.na(OS_days) & !is.na(EFS_days)) %>%
        dplyr::mutate(clk1_PSI = clk1_PSI * 10)
      
    }
  
    if (group == "DIPG or DMG"){
      group_hist <- group_hist %>%
       # dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
       dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, 
                                                 c("DMG, H3 K28", 
                                                   "DMG, H3 K28, TP53"))) %>%
        arrange(molecular_subtype)
    }
  
  # consolidate HGG subtypes into broader groups
  if (group == "Other high-grade glioma"){
    group_hist <- group_hist %>%
     # dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("HGG, H3 wildtype", "HGG, H3 wildtype, TP53",
                                                                "DHG, H3 G35", "DHG, H3 G35, TP53",
                                                                "HGG, IDH, TP53", "IHG, ALK-altered", 
                                                                "IHG, NTRK-altered", "IHG, ROS1-altered"
                                                                ))) %>%
      arrange(molecular_subtype)
  }
  
  if (group == "Low-grade glioma"){
    group_hist <- group_hist %>%
     # dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("Wildtype", "BRAF V600E", "BRAF fusion",
                                                                "Other alteration", "SEGA"
                                                                )))
  }
  
  if (group == "Ependymoma"){
    group_hist <- group_hist %>%
     # dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("EPN, PF A", "EPN, PF B",
                                                                "EPN, ST ZFTA", "EPN, ST YAP1",
                                                                "EPN, MPE", "EPN, SP", "EPN, SP-MYCN"
                                                                ))) %>%
   arrange(molecular_subtype)
  }
  
  if (group == "Mixed neuronal-glial tumor"){
    group_hist <- group_hist %>%
     # dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("Wildtype", "BRAF V600E", "BRAF fusion",
                                                                "Other alteration"
                                                                )))
  }
  
  if (group == "Medulloblastoma"){
    group_hist <- group_hist %>%
     # dplyr::mutate(mol_sub_group = factor(mol_sub_group)) %>% 
     dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group, c("MB, WNT", "MB, Group3",
                                                                "MB, Group4", "MB, SHH"
                                                                )))
    }
  
  # Run cox proprtional hazards models for OS, including appropriate covariates per histology
  add_model_os <- suppressWarnings(
                                fit_save_model(group_hist,
                                  terms = ifelse(
                                      group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                             "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                             "Medulloblastoma"),
                                      "mol_sub_group+extent_of_tumor_resection+clk1_PSI",
                                      ifelse(group %in% c("Ependymoma",
                                                          "DIPG or DMG"),
                                             "mol_sub_group+clk1_PSI",
                                              "clk1_PSI"
                                  )),
             coxph_os_model_file,
             "multivariate",
             years_col = "OS_years",
             status_col = "OS_status"
            ))
  
  if (!is.na(add_model_os$p.value[add_model_os$term == "clk1_PSI"])){
  
    add_os_forest_plot <- plotForest(readRDS(coxph_os_model_file))
  
    ggsave(file.path(plot_dir, glue::glue("forest_{plot_names[group]}_OS_add_subtype_clk1_psi.pdf")), 
           add_os_forest_plot, width = 7.5, height = 3)
    
  }
  
  # Run cox proprtional hazards models for PFS, including appropriate covariates per histology
  add_model_efs <- suppressWarnings(
                            fit_save_model(group_hist,
                                  terms = ifelse(
                                      group %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                             "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                             "Medulloblastoma"),
                                      "mol_sub_group+extent_of_tumor_resection+clk1_PSI",
                                      ifelse(group %in% c("Ependymoma",
                                                          "DIPG or DMG"),
                                             "mol_sub_group+clk1_PSI",
                                              "clk1_PSI"
                                  )),
             coxph_efs_model_file,
             "multivariate",
             years_col = "EFS_years",
             status_col = "EFS_status"
            ))
  
  if (!is.na(add_model_os$p.value[add_model_os$term == "clk1_PSI"])){
  
    add_efs_forest_plot <- plotForest(readRDS(coxph_efs_model_file))
  
    ggsave(file.path(plot_dir, glue::glue("forest_{plot_names[group]}_EFS_add_subtype_clk1_psi.pdf")), 
           add_efs_forest_plot, width = 7.5, height = 3)
    
  }
  
}
```


```{r}
# Define identifiers for model files
file_names <- c(glue::glue("{subtype_df$names}_{subtype_df$subtype_name}"))

# Loop through subtypes and run survival analyses
for (i in 1:nrow(subtype_df)) {
  
  results_dir <- file.path(analysis_dir, "results", subtype_df$foldername[i])
  
  plot_dir <- file.path(analysis_dir, "plots", subtype_df$foldername[i])
  
  # define output files

  coxph_os_model_file <- ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                                           "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                                           "Medulloblastoma"),
                             file.path(results_dir,
                                      glue::glue("cox_{file_names[i]}_OS_additive_terms_subtype_resection_clk1_psi.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{file_names[i]}_OS_additive_terms_subtype_clk1_psi.RDS"))
  )

  coxph_efs_model_file <- ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                                           "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                                           "Medulloblastoma"),
                             file.path(results_dir,
                                       glue::glue("cox_{file_names[i]}_EFS_additive_terms_subtype_resection_clk1_psi.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{file_names[i]}_EFS_additive_terms_subtype_clk1_psi.RDS"))
  )
  
    #Subset hist for histology subtype
    subtype_hist <- hist %>%
      filter(Histology == subtype_df$hist[i] & grepl(subtype_df$subtype[i], mol_sub_group)) %>%
      dplyr::mutate(clk1_PSI = clk1_PSI * 10)
    
    if (subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                 "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                 "Medulloblastoma")){

      subtype_hist <- subtype_hist %>%
        filter(!is.na(extent_of_tumor_resection) & !extent_of_tumor_resection %in% c("Unavailable", "Not Reported"))
      
    }


    # Run coxph models and save to output
    add_model_os <- suppressWarnings(
                              fit_save_model(subtype_hist,
                                  terms = ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                                                           "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                                                           "Medulloblastoma"),
                                                 "extent_of_tumor_resection+clk1_PSI",
                                                        "clk1_PSI"
                                                        ),
                                   coxph_os_model_file,
                                   "multivariate",
                                   years_col = "OS_years",
                                   status_col = "OS_status"))
    
    if (!is.na(add_model_os$p.value[add_model_os$term == "clk1_PSI"])){
    
      add_os_forest_plot <- plotForest(readRDS(coxph_os_model_file))
  
      ggsave(file.path(plot_dir, glue::glue("forest_{file_names[i]}_OS_add_subtype_clk1_psi.pdf")), 
             add_os_forest_plot, width = 7.5, height = 3)
      
    }
    
    add_model_efs <- suppressWarnings(
                              fit_save_model(subtype_hist,
                                  terms = ifelse(subtype_df$hist[i] %in% c("Mixed neuronal-glial tumor", "Low-grade glioma",
                                                                           "Other high-grade glioma", "Atypical Teratoid Rhabdoid Tumor",
                                                                           "Medulloblastoma"),
                                                 "extent_of_tumor_resection+clk1_PSI",
                                                        "clk1_PSI"
                                                        ),
                                     coxph_efs_model_file,
                                     "multivariate",
                                     years_col = "EFS_years",
                                     status_col = "EFS_status"
                                    ))
    
    if (!is.na(add_model_efs$p.value[add_model_efs$term == "clk1_PSI"])){
    
      add_efs_forest_plot <- plotForest(readRDS(coxph_efs_model_file))
  
      ggsave(file.path(plot_dir, glue::glue("forest_{file_names[i]}_EFS_add_subtype_clk1_psi.pdf")), 
             add_efs_forest_plot, width = 7.5, height = 3)
      
    }

}
```
