---
title: 'GSVA score comparisons in CLK1 treated cell lines'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Jo Lynne Rokita
date: "2024"
---

This script compares GSVA scores for hallmark and KEGG pathways between PBTA samples with and without DNA repair gene germline PLP variants

Load packages and set directories

```{r load libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(tibble)
library(ComplexHeatmap)
library(circlize)
```

Set filepaths
```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "CLK1-splicing-impact-morpholino")
results_dir <- file.path(analysis_dir, "results")
#source(file.path(analysis_dir, "util", "gsva_functions.R"))
source(file.path(root_dir, "figures", "theme_for_plots.R"))
```

Set file paths

```{r set file paths}
gsva_file <- file.path(results_dir, "clk1_ctrl_morpho_hallmark_gsva_scores.tsv")
kegg_file <- file.path(results_dir, "clk1_ctrl_morpho_kegg_gsva_scores.tsv")
dna_file <- file.path(results_dir, "clk1_ctrl_morpho_dna_repair_gsva_scores.tsv")
```

Create histology file for grouping samples

```{r}
sample_ids <- c("CTRL1", "CTRL2", "CTRL3", "Treated1", "Treated2", "Treated3")
treatments <- c("Control", "Control", "Control", "CLK1", "CLK1", "CLK1")
df <- tibble(sample_id = sample_ids, treatment = treatments)

# read in scores
hallmark_scores <- read_tsv(gsva_file)
hallmark_pathways <- unique(hallmark_scores$geneset)

kegg_scores <- read_tsv(kegg_file)
kegg_pathways <- unique(kegg_scores$geneset)

dna_scores <- read_tsv(dna_file)
dna_pathways <- unique(dna_scores$geneset)
```

Run GSVA score comparisons between samples with and without CLK1 morpholino

```{r}
# Define repair gene source input directories 
treat_ids <- df %>%
  filter(treatment == "CLK1") %>%
  pull(sample_id) %>%
  unique()

ctrl_ids <- df %>%
  filter(treatment == "Control") %>%
  pull(sample_id) %>%
  unique()
```

# define data frame to store GSVA scores, differences between groups, and corresponding test statistic p-values 
# Assuming you have defined kegg_pathways, hallmark_pathways, dna_pathways lists
# And kegg_scores, hallmark_scores, dna_scores data frames with columns 'geneset', 'sample_id', and 'gsva_score'
# treat_ids and ctrl_ids are vectors containing IDs for treated and control samples
```{r}
# Lists of score data frames and pathway lists
score_sources <- list(kegg_scores, hallmark_scores, dna_scores)
pathway_sources <- list(kegg_pathways, hallmark_pathways, dna_pathways)
score_names <- c("kegg", "hallmark", "dna_repair") # Names for output files

for (i in 1:length(score_sources)) {
  score_df <- score_sources[[i]]
  pathway_list <- pathway_sources[[i]]
  score_name <- score_names[i]
  
  gsva_res <- data.frame(pathway = pathway_list,
                         treat_score = rep(0, length(pathway_list)),
                         ctrl_score = rep(0, length(pathway_list)),
                         score_diff = rep(0, length(pathway_list)),
                         wilcox_stat = rep(0, length(pathway_list)),
                         pvalue = rep(0, length(pathway_list)))

 for (k in 1:length(pathway_list)) {
    treat_scores <- score_df[score_df$geneset == pathway_list[k] & score_df$sample_id %in% treat_ids, ]$gsva_score
    ctrl_scores <- score_df[score_df$geneset == pathway_list[k] & score_df$sample_id %in% ctrl_ids, ]$gsva_score
    
    # Ensure treat_scores and ctrl_scores are numeric vectors
    treat_scores <- as.numeric(treat_scores)
    ctrl_scores <- as.numeric(ctrl_scores)
    
    # Proceed with mean calculation and Wilcoxon test as before
    gsva_res$treat_score[k] <- mean(treat_scores, na.rm = TRUE)
    gsva_res$ctrl_score[k] <- mean(ctrl_scores, na.rm = TRUE)
    gsva_res$score_diff[k] <- gsva_res$treat_score[k] - gsva_res$ctrl_score[k]
    
    if (length(treat_scores) > 0 && length(ctrl_scores) > 0) {
      test_result <- tryCatch({
        wilcox.test(treat_scores, ctrl_scores, alternative = 'two.sided')
      }, warning = function(w) {
        return(list(statistic = NA, p.value = NA))
      }, error = function(e) {
        return(list(statistic = NA, p.value = NA))
      })
      
      gsva_res$wilcox_stat[k] <- test_result$statistic
      gsva_res$pvalue[k] <- test_result$p.value
    } else {
      gsva_res$wilcox_stat[k] <- NA
      gsva_res$pvalue[k] <- NA
    }
}

  
  # Write to output
  output_filename <- paste0("gsva_score_diff_", score_name, ".tsv")
  write_tsv(gsva_res, file.path(results_dir, output_filename))
}
```

Plot single samples
```{r}
# create anno colors
top_annot <- df %>%
  dplyr::select(treatment) %>%
  as.data.frame()
anno_col <- list(treatment = c("Control" = "#2166ac", "CLK1" = "#b2182b"))

# Heatmap annotation
top_anno = HeatmapAnnotation(df = top_annot,
                         col = anno_col, show_legend = TRUE)
 
# plot
for (i in 1:length(score_sources)) {
  score_df <- score_sources[[i]]
  score_name <- score_names[i]
  
  gsva_scores_mat <- score_df %>%
  spread(sample_id, gsva_score) %>%
  tibble::column_to_rownames("geneset") %>%
  as.matrix()
  
  if (i == 1){
    height <- 24
    width <- 24
  }
  else if (i == 2){
    height <- 12
    width <- 20
  }
  else if (i == 3){
    height <- 2
    width <- 6
  }
  
gsva_heatmap <- Heatmap(
  gsva_scores_mat,
  col = colorRamp2(c(-1, 0, 1), c("navy", "white", "orange")),
  name = "GSVA Scores",
  na_col = "whitesmoke",
  show_column_names = FALSE,
  cluster_columns = FALSE,
  row_names_gp = grid::gpar(fontsize = 12),
  top_annotation = top_anno,
  heatmap_legend_param = list(direction = "vertical")
)

output_filename <- paste0("gsva_heatmap_", score_name, ".pdf")
pdf(file.path(results_dir, output_filename),  height = height, width = width)
print(gsva_heatmap)
dev.off()
}

```

